---
title: "STAT8010 Mid term Assignment #1"
author: "Bernard McNamee"
date: "09/11/2020"
output: html_document
---

```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(xlsx)
library(kableExtra) # used here to format tables, eg highlight column with yellow background, 
library(gridExtra) # 
library(tidyverse)
```
# Introduction

This R statistical analysis report details the STAT8010 Mid term Assignment #1 with answers to the 11 questions asked. The report is generated using the RMarkdown package as a HTML document using libraries 'kableExtra' and 'gridExtra' for formatting and using libraries 'tidyverse' and 'xlsx' for data manipulation. The dataset is extracted from two sheets of an Excel spreadsheet titled 'assignment1.xlsx' into R. The questions posed require cleaning, transformation and analysis of the data with answers in text and plot/chart output.
\\

# Questions

**Q1. Using the xlsx or readxl package or otherwise, read each sheet in the ”assignment1.xlsx” file into R.**
```{r include=FALSE}
setwd("/home/barnyard/rstudio/scripts/R_CIT/Ass")
read.xlsx("assignment1.xlsx","Sheet1") # read xls file 1st sheet; head(Sheet1)
read.xlsx("assignment1.xlsx","Sheet2") # read xls file 2nd sheet; head(Sheet2)
```
The package 'read.xlsx' is installed and loaded into R and then using the library, the two sheets are separately read into R.  
\
\
**Q2. Generate a data frame for each sheet in the file.**

The process is repeated but this time the data is saved into two new dataframes, ie 'Sheet1' and 'Sheet2' are read into dataframes 'Sheet1' and 'Sheet2' respectively.  
\
A peek at the data reveals the structure and example values for each sheet.  
\
```{r size='tiny', echo=FALSE, fig.cap="Sheet 1 and Sheet 2 dataframes"}
Sheet1 <- read.xlsx("assignment1.xlsx","Sheet1", stringsAsFactors=FALSE) # read xls file 1st sheet into dataframe 'Sheet1' and read factors as characters/strings
Sheet2 <- read.xlsx("assignment1.xlsx","Sheet2", stringsAsFactors=FALSE) # read xls file 2nd sheet into dataframe 'Sheet2' and read factors as characters/strings
str(Sheet1)                                                       # take a peek at data
str(Sheet2)                                                       # take a peek at data
```
\
\
**Q3. The dataset in the first sheet is a random selection from a larger dataset. You will never get access to the full dataset so you should regenerate a new identification number for each subject in the dataset. This should be the row number of each entry in Sheet 1. You do not need to do this for Sheet 2.**  
\
A new identifier called 'newID' is created and assigned to each row of the dataframe - the value of each 'newID' is equal to the row number.  
\
A peek at the data reveals the head of the dataframe 'Sheet1' and the **new variable appears as the last column.**  
\
```{r echo=FALSE}
for (row in 1:nrow(Sheet1)) {                                                   # loop through each row of df 'Sheet1'
  Sheet1$newID[row] <- row                                                      # assign value of row number to new var 'newID'
}
head(Sheet1) %>%  kbl() %>%  kable_styling(font_size = 10) %>% column_spec(12, background = "yellow", bold=T)
```
\
\

**Q4. It is also required to have an additional identifier which is the number you have generated in (3) followed by the first letter of each subjects first name and then followed by the first letter of each subject’s surname. You do not need to do this for Sheet 2.**  
\
A new identifier called 'otherID' is created and assigned to each row of the dataframe - the value of each 'otherID' is equal to 'newID' + 1st letter First Name + 1st letter Surname.  
\
A peek at the data reveals the head of the dataframe 'Sheet1' and the **new variable appears as the last column.**  
\
```{r echo=FALSE, size="tiny"}
for (row in 1:nrow(Sheet1)) {                                                   # loop through each row of df 'Sheet1'
  # assign new value to new var 'otherID' - new value is a concatenation of row number, 1st char of 'FirstName' and 1st char of 'Surname'
  Sheet1$otherID[row] <- paste(row,substr(Sheet1$FirstName[row],1,1),substr(Sheet1$Surname[row],2,2), sep='')
}
head(Sheet1) %>%  kbl() %>%  kable_styling(font_size = 10) %>% column_spec(13, background = "yellow", bold=T)
```
\
\
**Q5. Although the data is not available for most subjects, some data highlighting subjects state of health is available in Sheet 2. You should use the subjects ID number to match it and merge it with the data in Sheet 1.**  
\
Health values in Sheet2 are found using two nested loops to cycle through each row of each sheet and check for available values and update Sheet1 with a new Health variable for all values found.  
\
A peek at the data reveals the head of the dataframe 'Sheet1' and the **new variable appears as the last column.**  
\
```{r echo=FALSE}
for (row1 in 1:nrow(Sheet1)) {                                                  # loop through each row of df 'Sheet1' 
  for (row2 in 1:nrow(Sheet2)) {                                                # loop through each row of df 'Sheet2'
    # evaluate True or False: if Sheet1 IDNumber matches Sheet2 IDNumber AND also if Sheet2 IDNumber value is available, ie not 'NA'
    if ((Sheet1$IDNumber[row1] == Sheet2$IDNumber[row2]) && !is.na(Sheet2$IDNumber[row2])) {
      # if previous statement True then this statement is executed - assign Sheet2 Health value to Sheet2 Health value
      Sheet1$Health[row1] <- Sheet2$Health[row2]                                
      break                                                                     # row found so no need to continue searching Sheet2 --> break out of loop
    } else {                                                                    # if the if statement not True then execute the following line
      Sheet1$Health[row1] <- 'NA'                                                # otherwise assign 0 to represent no health status
  }
}}
head(Sheet1) %>%  kbl() %>%  kable_styling(font_size = 10) %>% column_spec(14, background = "yellow", bold=T)
```
\
**Q6. Not every subject has its ID number included in Sheet 2. You should attempt to match the remaining subjects using their first and surnames. This must be done using tidyverse in a robust manner. Your code for doing this should work again in the case of a new sample of data being provided.**  
\
\
This is the filtered dataset for 'Sheet 2' with missing ID numbers.  
\
```{r echo=FALSE}
x <- Sheet2 %>% filter(is.na(IDNumber))                                         # create a new dataframe 'x' with only rows (filter) from Sheet2 that have IDNumber = 'NA'
x %>%  kbl() %>%  kable_styling(font_size = 10, full_width = F, position = "left") %>% column_spec(1, background = "yellow", bold = T)
```
\
This is the head of the new dataset for 'Sheet 1' following an update to find missing Health values.  
\
```{r include=TRUE,echo=FALSE}
# create a new dataframe 'y' with ;
# 1. rows (filter) from Sheet1 that have Health = 'NA'; 
# 2. rows (filter) that have Sheet1 FirstName = Sheet2 FirstName and Sheet1 Surname = Sheet2 Surname - and make vectors in each dataframe equal length
# 3. change (mutate) Sheet1 Health value to match Sheet2 Health value - if FirstName and Surname values also match
y <- Sheet1 %>% filter(Sheet1$Health == 'NA') %>% filter(FirstName %in% x$FirstName & Surname %in% x$Surname) %>% mutate(Health = ifelse(((FirstName==x$FirstName)&(Surname==x$Surname)), x$Health, Health))
Sheet1[match(y$IDNumber, Sheet1$IDNumber), ] <- y                               # overwrite rows in Sheet1 with all rows in dataframe y (with updated Health)
head(Sheet1) %>%  kbl() %>%  kable_styling(font_size = 10) %>% row_spec(6, background = "yellow", bold = T)
                                                               
```
_Note the Health for the subject on the last row (Hollis Byrne) has now been updated to **'2'** (from **'NA'** in table in previous question)._  
\
\
**Q7. You should add a column for age range. This should be**  
<table>
<tr><th style = 'padding:10px'><b>Age Range</b></th><th><b> Category Name</b></th></tr>
<tr><th>0-17</th><th>1</th></tr>
<tr><th>18-35</th><th>2</th></tr>
<tr><th>35-54</th><th>3</th></tr>
<tr><th>54-74</th><th>4</th></tr>
<tr><th>74-</th><th>5</th></tr>
</table>
\
\
A new variable 'Age Range' **appears in the last column** and corresponds to values in the table above.  
\
```{r include=TRUE,echo=FALSE}
Sheet1$AgeRange <- 'NA'                                                         # add new column titled 'Age Range' with no (NA) values
Sheet1$AgeRange[Sheet1$Age < 18] <- "1"                                         # assign all records with Age < 18 a label '1'
Sheet1$AgeRange[Sheet1$Age >= 18 & Sheet1$Age < 35] <- "2"                      # assign all records with Age 18 - 35  a label '2'
Sheet1$AgeRange[Sheet1$Age >= 35 & Sheet1$Age < 54] <- "3"                      # assign all records with Age 35 - 54 a label '3'
Sheet1$AgeRange[Sheet1$Age >= 54 & Sheet1$Age < 74] <- "4"                      # assign all records with Age 54 - 74 a label '4'
Sheet1$AgeRange[Sheet1$Age >= 74] <- "5"                                        # assign all records with Age > 74 a label '5'
head(Sheet1) %>%  kbl() %>%  kable_styling(font_size = 10) %>% column_spec(15, background = "yellow", bold = T)
```
\
\
**Q8. You should filter the data by each age category. Generate a bar plot using ggplot2 for the criminal record variable.**  
\
The two variables 'Age Range' and 'Criminal Record' are filtered into a dataframe, then a table created to record the frequency of 'Age Range' values for each level of 'Criminal Record', and finally a ggplot2 function histogram plot is created.  
```{r include=FALSE,echo=FALSE}
df <- Sheet1 %>% select(AgeRange, CriminalRecord)                               # select AgeRange and CriminalRecord from Sheet1 and then group by AgeRange
df <- data.frame(table(df))                                                     # create frequency table of values
label = 'Age Range\tCategory Name\n\t0-17\t\t\t1\n\t\t18-35\t\t2\n\t\t35-54\t\t3\n\t\t54-74\t\t4\n\t\t74-\t\t\t5'
gg <- ggplot(df, aes(AgeRange,Freq,fill=CriminalRecord)) + geom_histogram(stat='identity',bins = 30) + labs(x="Age Range (years)", y="Count") + geom_text(aes(label = label, x = -Inf, y = Inf),size=3, hjust = -2.1, vjust = 1.1)
```
```{r include=TRUE,echo=FALSE}
gg
```
\
\
\
**Q9. You should generate an appropriate visualisation examining the relationships between height, weight, age and criminal records. Comment on this.**  
\
The three variables are filtered into a new dataframe and each plotted against the other. These charts show the relationship between each variable and the other and appears to show a **null or no relationship** between any of the variables for this dataset.    
```{r include=TRUE,echo=FALSE}
df <- Sheet1 %>% select(Height.m., Weight.kg., Age, CriminalRecord)
plot(df)                                                                        
```
\
\
However, if we look more closely at each pair of variables, we see a different story.  
\
\
A **scatterplot with linear regression smoothing of 'Weight' v 'Height'** variables for each category of 'CriminalRecord (CR)' shows there is  
  - a **strong positive relationship** between 'Height' and 'Weight' for 'CR' = 1  
  - a **weak positive relationship** between 'Height' and 'Weight' for 'CR' = 2  
\
```{r include=FALSE,echo=FALSE}
CR=as.factor(Sheet1$CriminalRecord)                                             # change CriminalRecord to factor variable
gg <- ggplot(Sheet1, aes(Height.m.,Weight.kg., col=CR)) + geom_point() + facet_grid(~CriminalRecord) + geom_smooth(method=lm, formula = y ~ x) + labs(x="Height (m)", y="Weight (kg)")
```
```{r include=TRUE,echo=FALSE}
gg
```
\
\
\
For a different view of the data (and same variables), a **boxplot** shows  
  - the **'Weight' overall range, IQR range median is greater for CR=1** than CR=2    
\
```{r include=TRUE,echo=FALSE}
ggplot(Sheet1, aes(Height.m.,Weight.kg., col=CR)) + geom_boxplot() + facet_grid(~CriminalRecord) + labs(x="Height (m)", y="Weight (kg)")
```
\
\
Also, a **boxplot of 'Height' v 'Weight'** variables for each category of 'CriminalRecord (CR)' shows  
  - the **'Height' overall range, IQR range median is greater for CR=1** than CR=2  
\
```{r include=TRUE,echo=FALSE}
# boxplot for a different view of variable Height --> the 3rd quantile is smaller for CR=1 than CR=2 but overall range and median the same
ggplot(Sheet1, aes(Weight.kg.,Height.m., col=CR)) + geom_boxplot() + facet_grid(~CriminalRecord) + labs(y="Height (m)", x="Weight (kg)")
```
\
\
A scatterplot of **'Age' v 'Weight'** variables for each category of 'CriminalRecord (CR)' shows  
  - a **weak negative relationship** between 'Weight' and 'Age' when 'CR' = 1  
  - a **weak positive relationship** between 'Weight' and 'Age' when 'CR' = 2
\
```{r include=FALSE,echo=FALSE}
# boxplot for a different view of variable Height --> the 3rd quantile is smaller for CR=1 than CR=2 but overall range and median the same
gg <- ggplot(Sheet1, aes(Weight.kg.,Age, col=CR)) + geom_jitter() + facet_grid(~CriminalRecord) + geom_smooth(method=lm, formula = y ~ x) + labs(x="Weight (kg)", y="Age (years)")
```
```{r include=TRUE,echo=FALSE}
gg
```
\
\
And a scatterplot of **'Age' v 'Height'** variables for each category of 'CriminalRecord (CR)' shows  
  - a **weak negative relationship** between 'Height' and 'Age' when 'CR' = 1  
  - a **strong positive relationship** between 'Height' and 'Age' when 'CR' = 2
\
```{r include=FALSE,echo=FALSE}
# boxplot for a different view of variable Height --> the 3rd quantile is smaller for CR=1 than CR=2 but overall range and median the same
gg <- ggplot(Sheet1, aes(Height.m.,Age, col=CR)) + geom_jitter() + facet_grid(~CriminalRecord) + geom_smooth(method=lm, formula = y ~ x) + labs(x="Height (m)", y="Age (years)")
```
```{r include=TRUE,echo=FALSE}
gg
```
\
\
\
**Q10. Using filters, you should analyse if there are any interesting results in the dataset regarding the relationships between height, weight and criminal record. Use appropriate visualisations.**  
\
\
The dataset was filtered on 'Criminal Record' into two separate dataframes and then plotted as a scatterplot with linear regression and this plot shows  
  - a **weak positive relationship** between 'Weight' and 'Height' when 'CR' = 1  
  - a **strong positive relationship** between 'Weight' and 'Height' when 'CR' = 2  
\
```{r include=TRUE,echo=FALSE}
df1 <- Sheet1 %>% filter(CriminalRecord=='1') %>% select(Height.m.,Weight.kg.,CriminalRecord)
df2 <- Sheet1 %>% filter(CriminalRecord=='2') %>% select(Height.m.,Weight.kg.,CriminalRecord)
ggplot(df2,aes(Height.m.,Weight.kg.)) +  geom_point(aes(color = factor(CriminalRecord))) + geom_smooth(data=df1,method=lm, formula = y ~ x, aes(color = factor(CriminalRecord))) + geom_point(data=df1,aes(Height.m.,Weight.kg.,color = factor(CriminalRecord))) + geom_smooth(data=df2,method=lm, formula = y ~ x, aes(color = factor(CriminalRecord))) + labs(x="Height (m)", y="Weight (kg)", color = "Criminal Record")+ theme(legend.position = c(0.95, 0.95),legend.justification = c("right", "top"))
```
\
\
**Q11. Generate a smaller data frame for the subjects where health related data is available. Examine if there is a relationship between the different states of health and height, weight or age. Use appropriate visualisations. Note this should include a modelling type analysis such as regression. (S.Weisberg. Applied Linear Regression. Wiley Series in Probability and Statistics, 2005. may be useful)**  
\
\
A **violin plot of 'Health' v 'Height'** variables shows  
  - an uneven distribution of data points for both Health categories  
  - weak relationship; more tall (in range 1.8-2.2m) subjects with Health = 1  
  - weak relationship; more subjects with mid height than tall or short with Health = 2  
\
```{r include=TRUE,echo=FALSE}
df <- Sheet1 %>% select(Health, Height.m., Weight.kg., Age) %>% filter(Health != 'NA')
ggplot(df, aes(Health, Height.m., col=Health)) + geom_violin() + stat_summary(fun.data = mean_se, col="black") + labs(y="Height (m)")
```
\
\
A **violin plot of 'Health' v 'Weight'** variables shows  
  - an uneven distribution of data points for both Health categories  
  - weak relationship; more light (in range 60-75kg) subjects with Health = 1  
  - weak relationship; more subjects with mid weight than heavy or light with Health = 2  
\
```{r include=TRUE,echo=FALSE}
ggplot(df, aes(Health, Weight.kg., col=Health)) + geom_violin() + stat_summary(fun.data = mean_se, col="black") + labs(y="Weight (kg)")
```
\
\
A **violin plot of 'Health' v 'Age'** variables shows  
  - an uneven distribution of data points for both Health categories  
  - positive linear relationship; more old (in range 50-75years) subjects with Health = 1  
  - negative linear relationship; more young (in range 0-25) subjects with Health = 2
\
```{r include=TRUE,echo=FALSE}
ggplot(df, aes(Health, Age, col=Health)) + geom_violin() + stat_summary(fun.data = mean_se, col="black") + labs(y="Age (years)")
```
\
\
A **scatterplot of 'Height' v 'Weight'** variables with linear regression shows  
  - moderate positive linear relationship; ie taller subjects are heavier (no surprise there)
\
```{r include=TRUE,echo=FALSE}
ggplot(df, aes(Height.m.,Weight.kg.)) + geom_jitter() + geom_smooth(method=lm, formula = y ~ x) + labs(x="Height (m)", y="Weight (kg)")
```
\
\
A **scatterplot of 'Height' v 'Age'** variables with linear regression shows  
  - moderate positive linear relationship; ie taller subjects are older
\
```{r include=TRUE,echo=FALSE}
ggplot(df, aes(Height.m.,Age)) + geom_jitter() + geom_smooth(method=lm, formula = y ~ x) + labs(x="Height (m)", y="Age (years)")
```
\
\
A **scatterplot of 'Weight' v 'Age'** variables with linear regression shows  
  - null or no relationship
\
```{r include=TRUE,echo=FALSE}
ggplot(df, aes(Weight.kg.,Age)) + geom_jitter() + geom_smooth(method=lm, formula = y ~ x) + labs(x="Weight (kg)", y="Age (years)")
```
\
\

# Additional Interesting Findings

\
A histogram for each of the three variables shows the data is not uniform and may not reflect the general population at large.  
\
```{r include=TRUE,echo=FALSE}
p1 <- ggplot(Sheet1) + geom_histogram(mapping = aes(Height.m.),bins=15,binwidth=0.1)
p2 <- ggplot(Sheet1) + geom_histogram(mapping = aes(Weight.kg.),bins=15)
p3 <- ggplot(Sheet1) + geom_bar(mapping = aes(CR))
grid.arrange(p1, p2, p3, nrow = 1)
```
\
\
A histogram for 'each of the three variables'Education Level' shows the data that may not reflect the general population at large.  The assumption that most male (dataset has men only) criminals come from a lower socio-econonomic background with lower education levels **does not appear to be true for this dataset.**
\
```{r include=FALSE,echo=FALSE}
Sheet1Ed <- subset(Sheet1,(Sheet1$Education.Level!='FALSE'))                    # remove 'false' values, probably NAs
CR=as.factor(Sheet1Ed$CriminalRecord)                                           # change CriminalRecord to factor variable again
Sheet1Ed$Education.Level <- factor(Sheet1Ed$Education.Level,levels=c('Primary','secondary','graduate','masters','doctorate'),labels=c(A='1st',B='2nd',C='3rd',D='3M',E='PhD'))
gg <- ggplot(Sheet1Ed, aes(Education.Level, fill=CR)) + geom_histogram(stat='count',bins = 30) + facet_grid(~CriminalRecord)
```
```{r include=TRUE,echo=FALSE}
gg
```